Smooth risk-based predictive approaches to treatment effect heterogeneity: A simulation study

### 2021-06-16 submission/manuscript.rmd
first complete version of manuscript

In the simulated datasets based on the base-case scenario treatment was
allocated at random using a 50/50 split. For each patient we simulated $8$
baseline covariates, where $x_1,\dots,x_4\sim N(0, 1)$ and
$x_5,\dots,x_8\sim B(1, 0.2)$. Outcomes for patients in the control arm were
generated from a logistic regression model including all baseline covariates.
Coefficient values were such, so that the prediction model had an AUC of $0.75$
and an event rate of $20%$ in the control arm was achieved. Outcomes in the
treatment arm were created using the same logistic regression model, including a
constant treatment effect odds ratio (OR) of $0.8$. The generated samples of the
base-case scenario were of size $n=4,250$ ($80%$ power for the detection of an
unadjusted OR of $0.8$).

We evaluated the effect of sample size considering additional scenarios with
sample sizes of $1,064$ and $17,000$. We also evaluated the effect of prediction
performance, adjusting the baseline covariate coefficients, so that AUC values
of $0.65$ and $0.80$ were achieved when validating in a simulated dataset of
$500,000$ patients.

@@

When focusing on calibration for benefit, the linear interaction model had the
lowest median ICI for benefit in the majority of the scenarios except for the
scenarios where moderate linear deviations from the base case were considered.
In that case constant treatment effect models demonstrated the best performance,
very comparable to the linear interaction model's performance, nonetheless
(Figure \@ref(fig:calibration); Panels A, B, and C).

@@

We used the risk linear predictor to fit the the proposed methods under study
for individualizing absolute benefit predictions. All methods had quite
comparable results, in the sense that we predicted increasing benefits for
patients with higher baseline risk predictions. In terms of c-for benefit 
(validated internally) all models had quite comparable performance ranging from
0.519 (RCS smoothing with 3 knots) to 0.542 (RCS smoothing with 4 knots).
Similar conclusions could be drawn in terms of ICI-for-benefit which ranged
from 0.0039 (linear interaction approach) to 0.0053 (RCS smoothing with 4
knots).

The adaptive approach picked the
model with RCS smoothing with 4 knots, which had quite comparable performance to
the smooth fit with 5 knots. However, for lower baseline risk these 2 models
predicted implausible benefits and maybe should be avoided when applying such
models in practice. The linear interaction model and the model with RCS
smoothing (3 knots) made very similar predictions, while also followed quite
closely the evolution of the stratified estimates. In this case, we observed
that even using a simple model with a constant relative treatment effect can
result in very similar benefit predictions as other more complex approaches.

@@

# Discussion

>>>>

In the simulated datasets based on the base case scenario treatment was
allocated at random using a 50/50 split. For each patient we simulated $8$
baseline covariates, where $x_1,\dots,x_4\sim N(0, 1)$ and
$x_5,\dots,x_8\sim B(1, 0.2)$. Outcomes for patients in the control arm were
generated from a logistic regression model including all baseline covariates.
Coefficient values were such, so that the prediction model had an AUC of $0.75$
and an event rate of $20\%$ in the control arm was achieved. Outcomes in the
treatment arm were created using the same logistic regression model, including a
constant treatment effect odds ratio (OR) of $0.8$. The generated samples of the
base case scenario were of size $n=4,250$ ($80\%$ power for the detection of an
unadjusted OR of $0.8$).

We evaluated the effect of sample size considering additional scenarios with
sample sizes of `r 1064` and `r 17000`. We also evaluated the effect of prediction
performance, adjusting the baseline covariate coefficients, so that AUC values
of $0.65$ and $0.80$ were achieved when validating in a simulated dataset of
$500,000$ patients.

@@

When focusing on calibration, the linear interaction model had the lowest median
ICI for benefit in the majority of scenarios except for the scenarios where
moderate linear deviations from the base case were considered. In that case
constant treatment effect models demonstrated the best performance, very
comparable to the linear interaction model's performance, nonetheless (Figure
\@ref(fig:calibration); Panels A, B, and C).

@@

We used the risk linear predictor to fit the the proposed methods under study
for individualizing absolute benefit predictions. All methods had quite
comparable results, in the sense that we predicted increasing benefits for
patients with higher baseline risk predictions. In terms of c-for-benefit 
(validated internally) all models had quite comparable performance ranging from
0.519 (RCS smoothing with 3 knots) to 0.542 (RCS smoothing with 4 knots).
Similar conclusions could be drawn in terms of ICI-for-benefit which ranged
from 0.0039 (linear interaction approach) to 0.0053 (RCS smoothing with 4
knots).

The adaptive approach picked the
model with RCS smoothing with 4 knots, which had quite comparable performance to
the smooth fit with 5 knots. However, for lower baseline risk these 2 models
predicted implausible benefits and maybe should be avoided when applying such
models in practice. The linear interaction model and the model with RCS
smoothing (3 knots) made very similar predictions and also followed quite
closely the evolution of the stratified estimates. In this case, we observed
that even using a simple model with a constant relative treatment effect can
result in very similar benefit predictions as other more complex approaches.

@@

# Discussion
We have carried out a simulation study comparing different approaches for
individualized benefit prediction using baseline risk as the reference. These
methods included a constant treatment effect model, a model including a simple
linear interaction of the risk linear predictor with treatment, and models using
smoothing with restricted cubic splines with 3, 4, and 5 knots. Performance was
assessed evaluating error rates using RMSE and benefit prediction performance
using some novel methods for evaluating discrimination and calibration for
benefit.

The linear interaction model proved to be flexible enough to adequately predict
benefit under a wide range of scenarios. It had lower error rates and usually
the highest c-for-benefit with the lower variability across simulation
replications. Its benefit predictions were also well-calibrated across the
majority of the simulation settings. Models with RCS smoothing with 3 knots
performed very similarly, only outperformig the linear interaction model in the
presence of strong quadratic deviations from the base case scenario of constant
relative treatment effects. These simpler methods proved to have quite robust
performance across all scenarios, which was not the case for more flexible
approaches.

Methods with RCS smoothing with 4 or 5 knots proved to be data-hungry and this
often led to overfitting with smaller sample sizes. This resulted in higher
RMSE, increased variability of discrimination for benefit and worse-calibrated
benefit predictions across simulation replications. Even with larger sample
sizes and strong quadratic deviations from the base case scenario of constant
relative treatment effects, these more flexible RCS smoothing methods did not
outperform simpler RCS smoothing with 3 knots. More flexible approaches usually
do not come with any added value, unless we expect extreme heterogeneity in
treatment effects.

The constant treatment effect model, despite having adequate performance even in
the presence of weak treatment effect heterogeneity on the relative scale, it
quickly broke down with stronger deviations from the base case scenario of
constant effects. In these cases, the stratified approach generally had lower
error rates compared to the constant treatment effect model. In concept, the
stratified approach lies between the constant effects model and smoother
approaches, only assuming constant treatment effects within strata of predicted
risk and, therefore, is more sensitive to treatment effect heterogeneity.
However, these stepwise benefit estimates can only be used as a demonstration of
treatment effect heterogeneity, identifying patient subgroups where treatment
benefits are expected to be higher or lower, and should not be used to make
individualized absolute benefit predictions.

Increasing the true prediction AUC by increasing the predictor coefficients of
the true risk model reduced RMSE for all methods, similar to increasing the
sample size. Higher AUC translates in higher variability of predicted risks,
which, in turn, allows the considered methods to better follow the evolution of
treatment benefit. Higher prediction AUC also led to higher absolute values of
c-for-benefit. Even though, model performance appears to be very important for
the performance of these methods, prediction model development was outside the
scope of this work and has already been studied quite extensively
[@vanKlaveren2019; @Burke2014; @Abadie2018].

Risk-based approaches to predictive HTE assume that benefit is a function of
baseline risk. A limitation of our study is that this was assumed to be the case
in the majority of the simulation scenarios. Even though our main conclusions
did not change when simulating true treatment-covariate interactions (results in
supplement), RMSE had higher values. The linear interaction model under these
scenarios again had the best overall performance under different sizes of the
interaction effects. This can be explained by the fact that linear interactions
with treatment were simulated, thus favoring the linear interaction model. In
studies evaluating prediction model development, the underlying outcome
generation process sometimes proved to affect performance of certain methods
[@Austin2021]. This might be the case here as well and should be explored in
more extensive future simulation studies.

Recent years have seen an increased interest in predictive HTE approaches
focusing on individualized benefit predictions. In our simulations we only
focused on risk-based methods, using baseline risk as a reference in a two-stage
approach to individualizing benefit predictions. However, there is a plethora of
different methods, ranging from treatment effect modeling to tree-based
approaches available in more recent literature [@Athey2019; @Lu2018;
@Wager2018]. Simulations are also needed to assess relative performance and
define the settings where these break down or outperform each other.

In conclusion, we carried out a simulation study comparing different risk-based
approaches to individualizing benefit predictions. Simpler models including a
linear interaction with the risk linear predictor or RCS smoothing with 3 knots
had adequate performance under a wide range scenarios. An adaptive approach based
on selecting among the considered methods using AIC had satisfactory
performance.
